Measuring contrast processing in the visual system using SSVEP

### 2024-05-06 review.qmd
More

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. For example, @Tsai2012 demonstrated that a gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. Our own work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

@@

# Attention

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006, @Muller1998m @MullerHillyard2000, MorganPNAS1996].

@@

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [Middendorf2000, ]. EEG signals are highly sensitive to both visual field position [@Ales2010] and attentional state [@Verghese2012; @Lauritzen2010], so the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties : For example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli to increase SNR or the range of stimuli that can be frequency-tagged (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....).

Probably a summary of BCIs here.

# Conclusions

>>>>

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. For example, @Tsai2012 demonstrated that a gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. Our own work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

@@

# Attention and arousal

Attention exerts a profound influence on visual performance: for example, allowing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006, @Muller1998m @MullerHillyard2000, @MorganPNAS1996, @Hillyard1998NNS].

@@

The SSVEP can also be used to study changes in visual processing \textit{driven} by different behavioural states or overall arousal. For example, locomotion has been shown to alter neuronal excitability and spatial normalization in mice [@NeilStryker2010, @Ayaz2013] - running mice have higher visual sensitivity and lower surround suppression compared to stationary mice. Although measuring EEG responses from locomoting humans is technically challenging, SSVEP studies (which are able to distinguish broadband noise from input signal effectively) have shown that walking also alters early visual processing although in a manner different to that observed in mice [@Benjamin2018, @CaoPlosBiol2019]

SSVEP measurements are typically used to measure time invariant responses due to sustained attention. However, recent work has shown that moderately high modulation frequencies (ca 10Hz) and short analysis windows (1.5s) can also be used to track the dyanmic allocation of attention across a task [@ChotaJCN2024] or attention to moving targets [@deLissaPlos12020]. The ability to track changes in attention over short time periods is also important if SSVEP is to be used for dynamic readouts - for example in a Brain Computer Interface.

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf]. BEcause SSVEP signals are highly sensitive to both visual field position [@Ales2010, @dirusso2007_HBM_spatiotemporal] and attentional state [@hillyard1996_PNAS, @Hillard1998_NNS, @Verghese2012; @Lauritzen2010],  the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is typically present, this modulation and the associated changes in gain control, will affect the entire pattern of self- and intermodulation terms allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties : For example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Conclusions
The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants) and at frequencies high enough to be effectively invisible [REF - Some high frequency work... - Romanian group]. The SSVEP is also able 'fingerprint' modulators of the input frequency through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by

### 2024-05-06 review.tex
More

Attention exerts a profound influence on visual performance: for
example, allowing people to attend covertly to a spatial location for
example, improves their performance on a target detection task
significantly (Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco
\emph{et al.}, 2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.},
2004; Pestilli \emph{et al.}, 2009). In principle, this enhancement may
be driven both by modulation of the underlying signal or noise
characteristics, or by additional decision-theoretic factors such as
reduction in spatial uncertainty (Petrov \emph{et al.}, 2006; Gould
\emph{et al.}, 2007). Early experiments with non-human primates showed
little evidence for attentionally-driven changes in neuronal spike rates
(Luck \emph{et al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et
al.}, 2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with
the advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI BOLD responses (Tootell \emph{et al.}, 1998; Somers
\emph{et al.}, 1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe,
1999; Kastner \emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver
\emph{et al.}, 2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG
signals (Ding \emph{et al.}, 2006).

Electrophysiological and psychophysical measurements of attentional
effects on both luminance and chromatic contrast strongly implicated
gain control an an underlying mechanism (Lu \& Dosher, 1998; Di Russo \&
Spinelli, 1999\emph{a}; Di Russo \emph{et al.}, 2001\emph{b}) and
demonstrated that these effects differed between chromatic and
achromatic pathways (Di Russo \& Spinelli, 1999\emph{b}). In the late
2000s a comprehensive theoretical model for attentional modulation was
developed that framed it as a gain control computation (Reynolds \&
Heeger, 2009; Boynton, 2009). This framework has proven to be
influential - explaining a wide range of phenomena from the earlier
literature and demonstrating subtle interactions between the size of the
attentional `spotlight' and the stimulus configuration which rationalise
many apparent contradictions in the literature. Subsequent papers
(effect of stimulus configuration critical). Then\ldots{} talk about
Lauritzen \emph{et al.} (2010) Verghese \emph{et al.} (2012) (1 other
Verghese paper) maybe mention Highsmith \& Crognale (2010), Wang \& Wade
(2011) and Martinovic \& Andersen (2018) Probably lots more Carasco
\ldots{}

@@

One widespread recent application of the SSVEP technique is in the
design of brain-computer interfaces, which seek to control some aspect
of a computer using neural signals. The high SNR and frequency
resolution of SSVEPs make them an ideal candidate for this approach.
Typical studies may involve presenting an array of stimuli at different
flicker frequencies, and having the participant select one either by
overt attention (i.e.~shifting fixation to foveate the selected
stimulus) or covert attention (i.e.~deploying attention to one stimulus
whilst keeping fixated). EEG signals are highly sensitive to both visual
field position (Ales \emph{et al.}, 2010) and attentional state
(Lauritzen \emph{et al.}, 2010; Verghese \emph{et al.}, 2012), so the
response to the selected stimulus will typically increase relative to
the others, allowing it to be identified by an on-line algorithm.

Probably a summary of BCIs here.

@@

@@

@@

@@

@@

>>>>

Attention exerts a profound influence on visual performance: for
example, allowing people to attend covertly to a spatial location
improves their performance on a target detection task significantly
(Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco \emph{et al.},
2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.}, 2004; Pestilli
\emph{et al.}, 2009). In principle, this enhancement may be driven both
by modulation of the underlying signal or noise characteristics, or by
additional decision-theoretic factors such as reduction in spatial
uncertainty (Petrov \emph{et al.}, 2006; Gould \emph{et al.}, 2007).
Early experiments with non-human primates showed little evidence for
attentionally-driven changes in neuronal spike rates (Luck \emph{et
al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et al.},
2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with the
advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI BOLD responses (Tootell \emph{et al.}, 1998; Somers
\emph{et al.}, 1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe,
1999; Kastner \emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver
\emph{et al.}, 2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG
signals (\textbf{MullerHillyard2000?}).

Electrophysiological and psychophysical measurements of the effect of
attention on both luminance and chromatic contrast have strongly
implicated gain control as an underlying mechanism (Lu \& Dosher, 1998;
Di Russo \& Spinelli, 1999\emph{a}; Di Russo \emph{et al.},
2001\emph{b}). These effects also appear to differ between chromatic and
achromatic pathways (Di Russo \& Spinelli, 1999\emph{b}) - perhaps as a
result of the different levels of nonlinear gain control in the early
pre-cortical magno-, parvo- and konio-cellular pathways (Derrington \&
Lennie, 1984; Kaplan \& Shapley, 1986; Lee \emph{et al.}, 1990; Solomon
\& Lennie, 2005). In the late 2000s a comprehensive theoretical model
for attentional modulation was developed that framed it as a gain
control computation (Reynolds \& Heeger, 2009; Boynton, 2009). This
framework has proven to be influential - explaining a wide range of
phenomena from the earlier literature and demonstrating subtle
interactions between the size of the attentional `spotlight' and the
stimulus configuration which rationalise many apparent contradictions in
the literature. Direct measurements of attentional modulation of
achromatic SSVEP signals are broadly consistent with this model
Martinovic \& Andersen (2018) and confirm the relatively weaker role of
spatial attention on responses driven by chromatic stimuli -
particularly those that isolate the opponent S-(L+M) cone pathway
Highsmith \& Crognale (2010). The gain control model of attention can
also be extended to SSVEP studies of feature-based attention which show
that the modulatory effects can be targeted to the most informative
neuronal populations (Verghese \emph{et al.}, 2012).

@@

One widespread recent application of the SSVEP technique is in the
design of brain-computer interfaces, which seek to control some aspect
of a computer using neural signals. The high SNR and frequency
resolution of SSVEPs make them an ideal candidate for this approach.
Typical studies may involve presenting an array of stimuli at different
contrast flicker frequencies, and having the participant select one
either by overt attention (i.e.~shifting fixation to foveate the
selected stimulus) or covert attention (i.e.~deploying attention to one
stimulus whilst keeping fixated) (\textbf{Middendorf?}). BEcause EEG
signals are highly sensitive to both visual field position (Ales
\emph{et al.}, 2010) and attentional state (Lauritzen \emph{et al.},
2010; Verghese \emph{et al.}, 2012), the response to the selected
stimulus will typically increase relative to the others, allowing it to
be identified by an on-line algorithm. Because more than one stimulus
frequency is typically present, this modulation and the associated
changes in gain control, will affect the entire pattern of self- and
intermodulation terms.

This approach is primarily useful in situations that require the BCI to
distinguish from among a small set of possibilties : For example, in
early work, visual stimuli representing `Left' and `Right' commands in a
flight simulator were distinguished robustly (\textbf{Middendorf?}).
Although SSVEP-based BCI interfaces typically do use contrast flicker,
work in this field has largely focused on optimising the stimuli or
decoders to increase decoding performance (e.g.~\ldots{} ) and reducing
visual fatigue associated with the long-term presentation of arrays of
high-contrast flicker (e.g.~\ldots..) rather than studying visual
contrast processing \textit{per se}.

@@

\bibitem[\citeproctext]{ref-Derrington1984b}
Derrington AM \& Lennie P (1984).
\href{https://doi.org/10.1113/jphysiol.1984.sp015498}{Spatial and
temporal contrast sensitivities of neurones in lateral geniculate
nucleus of macaque}. \emph{J Physiol} \textbf{357,} 219--240.

@@

\bibitem[\citeproctext]{ref-Hou2016}
Hou C, Kim Y-J, Lai XJ \& Verghese P (2016).
\href{https://doi.org/10.1167/16.3.16}{Degraded attentional modulation
of cortical neural populations in strabismic amblyopia}. \emph{Journal
of Vision} \textbf{16,} 16.

@@

\bibitem[\citeproctext]{ref-Kaplan1986}
Kaplan E \& Shapley RM (1986).
\href{https://doi.org/10.1073/pnas.83.8.2755}{The primate retina
contains two types of ganglion cells, with high and low contrast
sensitivity}. \emph{Proc Natl Acad Sci U S A} \textbf{83,} 2755--2757.

@@

\bibitem[\citeproctext]{ref-Lee1990}
Lee BB, Pokorny J, Smith VC, Martin PR \& Valberg A (1990).
\href{https://doi.org/10.1364/josaa.7.002223}{Luminance and chromatic
modulation sensitivity of macaque ganglion cells and human observers}.
\emph{J Opt Soc Am A} \textbf{7,} 2223--2236.

@@

\bibitem[\citeproctext]{ref-Solomon2005}
Solomon SG \& Lennie P (2005).
\href{https://doi.org/10.1523/JNEUROSCI.5316-04.2005}{Chromatic gain
controls in visual cortical neurons}. \emph{J Neurosci} \textbf{25,}
4779--4792.
